Prosper Loan Analysis by Anushree Mehra
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(dplyr)
library(gridExtra)
library(alr3)
library(reshape2)
library(GGally)
library(scales)
library(RColorBrewer)
library(grid)
```

# Load and modify dataset
```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
loandata <- read.table('C:/Users/anmehra/Data Analysis with R/prosperLoanData.csv', sep=",", header = TRUE)

# Understand the structure of dataset
str(loandata)
summary(loandata)


# Change the column names to avoid spceial characters not recognized by R
colnames(loandata)[colnames(loandata)=="ProsperRating..numeric."] <-
  "ProsperRating_numeric"
colnames(loandata)[colnames(loandata)=="ProsperRating..Alpha."] <-
  "ProsperRating_alpha"
colnames(loandata)[colnames(loandata)=="ListingCategory..numeric."] <-
  "ListingCategory_numeric"
colnames(loandata)[colnames(loandata)=="TradesNeverDelinquent..percentage."] <-
  "TradesNeverDelinquent_percentage"

# Subset the data to include the records from July 2009, start of the prosper rating process.
loandata_2009 <- subset(loandata, 
                        !is.na(ProsperRating_numeric))

# Add a new variable for mean credit score
loandata_2009$MeanCreditScore <- (loandata_2009$CreditScoreRangeLower +
                                    loandata_2009$CreditScoreRangeUpper)/2
```
I modified the dataset to only include the records with Prosper Rating, i.e. since July 2009. I also added a new variable MeanCreditScore, calculated from CreditScoreLowerRange and CreditScoreUpperRange, so that I can use it for comparisons and investigation.

# Univariate Plots Section

### Explore key variables for the new dataset. I have selected Prosper Rating, Estimated Loss Rate, Borrower Rate, and Lender Yield as main features of interest.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot1}

## Reorder ProsperRating_alpha 
loandata_2009$ProsperRating_alpha <- factor(loandata_2009$ProsperRating_alpha,
                                            levels = c("AA", "A", "B", "C", "D",
                                                       "E", "HR"), 
                                            ordered = TRUE)

## Distribution for Prosper Rating
ggplot(aes(x = ProsperRating_alpha), 
       data = loandata_2009) +
  geom_histogram(fill = 'blue') 

summary(loandata_2009$ProsperRating_alpha)

```
I reordered the ProsperRating_alpha factors, to display them in the sequential order in the charts. Prosper Rating has a normal distribution, with C category having the highest number of loans. Total number of loans for C category are 18345. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot2}

## Distribution for Estimated Loss Rate
ggplot(aes(x = EstimatedLoss), 
       data = loandata_2009) +
  geom_histogram(binwidth = 0.005) +
  xlim(0.0, 0.2)

summary(loandata_2009$EstimatedLoss)
```
We can see that there are few outliers in the chart after 0.2. We can also validate it looking at the statistics. 3rd quartile is at 0.112 and max is at 0.366. I modified the chart to limit the x axis and get a better idea of the distribution. The distribution looks right skewed with few exceptional peaks. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot3}

## Distribution for Borrower Rate
ggplot(aes(x = BorrowerRate), 
       data = loandata_2009) +
  geom_histogram(binwidth = 0.01)  +
  scale_x_continuous(breaks = seq(0, 0.4, 0.05))

summary(loandata_2009$BorrowerRate)

```
The distribution for borrower rate is mostly normal except a peak around 0.31. The mean borrower rate is 0.196 and the minimum is at 0.04. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot4}

## Distribution for Lender Yield
ggplot(aes(x = LenderYield), 
       data = loandata_2009) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(breaks = seq(0, 0.4, 0.05))

summary(loandata_2009$LenderYield)
```
The distribution is similar to borrower rate with a peak around 0.31. The mean lender yield is 0.178 and the minimum is at 0.03. 


### Explore other features of interest like MeanCreditScore, IncomeRange, LoanStatus, and EmploymentStatus.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot5}

## Distribution for Mean Creidt Score
ggplot(aes(x = MeanCreditScore), 
       data = loandata_2009) +
  geom_histogram() +
  scale_x_continuous(breaks = seq(600, 900, 25))

summary(loandata_2009$MeanCreditScore)
```
The distribution looks a little postively skewed. There is an unusual gap between 685 and 715, and then again between 830 and 850. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot6}

## Reorder IncomeRange
loandata_2009$IncomeRange <- factor(loandata_2009$IncomeRange, 
                                    levels = c("Not employed", "$0",
                                               "$1-24,999","$25,000-49,999",
                                               "$50,000-74,999",
                                               "$75,000-99,999", "$100,000+"),
                                    ordered = TRUE)


## Distribution of Income Range 
ggplot(aes(x = IncomeRange), 
       data = loandata_2009) +
  geom_histogram() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=40, hjust = 1, vjust= 1))

summary(loandata_2009$IncomeRange)

```
I reordered the IncomeRange factors, to display them in the sequential order in the charts.Distribution looks normal. People with $50,000-74,999 income range request the most number of loans i.e. 24175.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot7}

## Distribution of Loan Status 
ggplot(aes(x = LoanStatus), 
       data = loandata_2009) +
  geom_histogram() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=60, hjust = 1, vjust= 1))

summary(loandata_2009$LoanStatus)

```
Distribution is positively skewed. Most of the loans are either completed(repaid in full) or currently in process of repayment. There are very few loans that are either charged off, defaulted or past due dates. Total loans are 84853, Out of which 19664 loans are completed, 56576 loans are in current status, and 5336 are charged off. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot8}

## Distribution of Employment Status 
ggplot(aes(x = EmploymentStatus), 
       data = loandata_2009) +
  geom_histogram() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=40, hjust = 1, vjust= 1))

summary(loandata_2009$EmploymentStatus)

```
Distribution is positively skewed. Most of the borrowers are employed. Few have categorized themselves as full-time, half-time or self-employed. There are fewer who are not employed or retired. Total borrowers are 84853, out of which 67310 are employed, and only 649 are not employed. There are few more borrowers in other categories. Check the summary.

### Explore few other variables to understand general trends about the data, like number of loans by quarter/year, breakdown by state, etc. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot9}

## Reorder LoanOriginationQuarter
loandata_2009$LoanOriginationQuarter <- factor(loandata_2009$LoanOriginationQuarter, levels = 
                                                 c("Q3 2009", "Q4 2009", "Q1
                                                   2010", "Q2 2010", "Q3 2010",
                                                   "Q4 2010", "Q1 2011", "Q2
                                                   2011", "Q3 2011", "Q4 2011",
                                                   "Q1 2012", "Q2 2012", "Q3
                                                   2012", "Q4 2012", "Q1 2013",
                                                   "Q2 2013", "Q3 2013", "Q4
                                                   2013", "Q1 2014"), 
                                               ordered = TRUE)


## Frequency plot for total number of loans by quarter-year.
ggplot(aes(x = LoanOriginationQuarter), 
       data = subset(loandata_2009, !is.na(LoanOriginationQuarter))) + 
  geom_freqpoly(aes(group = 1)) +
  scale_y_continuous(breaks = seq(0, 30000, 5000)) +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=90, hjust = 1, vjust= 1))

```

I reordered the LoanOriginationQuarter factors, to display them in the sequential order in the charts. The frequency plot for total number of loans depicts that prosper loans are becoming popular with time. There is a steep rise from Q1 2013 to Q4 2013. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot10}
# Breakdown by State

## Groupby BorrowerState
loandata_2009_state_group <- group_by(loandata_2009, 
                                      BorrowerState)

## Get the count by BorrowerState
loandata_2009_by_state <- summarise(loandata_2009_state_group,
                          n = n())

## Get the states with greater than or equal to 3rd quartile of total loans
state_count_top25 <- subset(loandata_2009_by_state, 
                            n >= quantile(n, 0.75))

## View the top states, along with number of loans
ggplot(aes(x = BorrowerState, y = n), 
       data = state_count_top25) +
  geom_histogram(stat = 'identity') 

```

Here I am trying to get the top borrower states constituting of 75% of the total loans. As per my analysis, I have found top 12 borrower states and the histogram depicst the breakdown of number of loans. California is the top borrower state with 10761 loans.


# Univariate Analysis

### What is the structure of your dataset?
Prosper is America's first peer-to-peer lending marketplace, founded in 2005. Borrower can request personal loans, and propser verifies the borrower's identity and other related information before granting the loan. Prosper handles the servicing of the loan, collects borrower payments, and distributes interest back to the loan investors.

Prosper dataset has 81 variables. It contains the data since 2006 to early 2014, that constitute of 113937 observations. 

### What is/are the main feature(s) of interest in your dataset?
The main features are Prosper Rating, Estimated Loss Rate, Borrower Rate, and Lender Yield. From 2006 to 2009, Prosper had a variable rate model. From July 2009, Prosper started assigning Prosper Rating to prospective borrowers. According to the company, Prosper Rating is calculated based on borrower's estimated loss rate. 

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
Few other features that will be helpful to support the investigation are DebtToIncomeRatio, Income Range, Employment Status, Loan Status, Loan Original Amount, Investors, etc. 

### Did you create any new variables from existing variables in the dataset?
I modified the dataset to only include the records with Prosper Rating, i.e. since July 2009. I also added a new variable MeanCreditScore, calculated from CreditScoreLowerRange and CreditScoreUpperRange, so that I can use it for comparisons and investigation. 

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
- Distribution for BorrowerRate was mostly normal, except a peak around 0.31
- Distribution for ProsperRating was normal. 
- The distribution looks a little postively skewed. There is an unusual gap between 685 and 715, and then again between 830 and 850.
- The frequency plot for total number of loans depicts that prosper loans are becoming popular with time. There is a steep rise from Q1 2013 to Q4 2013. 
- I reordered the ProsperRating and LoanOriginationQuarter factors, to display them in the sequential order in the charts. 
- I also used the group by and summarise function to find the states with the highest number of loans.


# Bivariate Plots Section

### Explore relationship among main features.
```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot1}

# Relationship between EstimatedLoss and ProsperRating

## Scatterplot for EstimatedLoss vs ProsperRating. Omit the top 1%
ggplot(aes(x = ProsperRating_alpha, y = EstimatedLoss), 
       data = loandata_2009) +
  geom_point() +
  scale_y_continuous(lim = c(0, quantile(loandata_2009$EstimatedLoss, 
                                         0.99)))

## Find the correlation using Pearson's method
with(loandata_2009, cor.test(ProsperRating_numeric, EstimatedLoss))

## Use box plots to explore the min, max and median value of EstimatedLoss for each ProsperRating. Omit the top 1%
ggplot(aes(x = ProsperRating_alpha, y = EstimatedLoss), 
       data = loandata_2009) +
  geom_boxplot() +
  scale_y_continuous(lim = c(0, quantile(loandata_2009$EstimatedLoss, 
                                         0.99)))

## Get the statistics for EstimatedLoss by ProsperRating
by(loandata_2009$EstimatedLoss, 
   loandata_2009$ProsperRating_alpha, 
   summary)

```
This chart depicts the relatioship between estimated loss and prosper rating. I can see a strong correlation between estimated loss and prosper rating. To avoid the outliers, I am excluding the top 1% y-values. As calculated by Pearson's method, correlation is -0.96, with a 95% confidence interval. So, there is a strong negative correlation between both variables. Due to overlapping values, using box plot to better understand the statistics. Also calculated the statistics for each prosper rating. As we can see, median (0.015) for AA is quite close to Max value(0.0199). We can also see the evidence of outliers for HR, with Max being at 0.37 and 3rd quartile at 0.17. Max is way high when compared to 3rd quartile. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot2}

# Relationship between EstimatedReturn and ProsperRating

## Boxplot between EstimatedReturn and ProsperRating
ggplot(aes(x = ProsperRating_alpha, y = EstimatedReturn), 
       data = loandata_2009) +
  geom_boxplot() +
  scale_y_continuous(lim = c(0, quantile(loandata_2009$EstimatedReturn, 
                                         0.90)))

## Find the correlation using Pearson's method
with(loandata_2009, cor.test(ProsperRating_numeric, EstimatedReturn))

```
Due to many overlapping values, I decided to go with the box plot. I can see many outliers even after excluding top 10% values. Unlike Estimated Loss chart, the first quartile for one of the lower prosper rating might be less than the third quartile of the higher propser rating. I don't see a strong correlation between the two variables. To validate my findings, I calculated correlation by Pearson's method. correlation is -0.66, 95% confidence interval. So, they are related but do not have a strong relationship. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot3}

# Relationship between EstimatedEffectiveYield and ProsperRating

## Boxplot between EstimatedEffectiveYield and ProsperRating
ggplot(aes(x = ProsperRating_alpha, y = EstimatedEffectiveYield), 
       data = loandata_2009) +
  geom_boxplot() +
  scale_y_continuous(lim = c(0, quantile(loandata_2009$EstimatedEffectiveYield, 
                                         0.95)))

## Find the correlation using Pearson's method
with(loandata_2009, cor.test(ProsperRating_numeric, EstimatedEffectiveYield))

## Get the statistics for EstimatedEffectiveYield by ProsperRating
by(loandata_2009$EstimatedEffectiveYield, 
   loandata_2009$ProsperRating_alpha, 
   summary)

```
I can see many outliers even after excluding top 5% values. Like Estimated Loss chart, the quartile boundaries are separated for all the propsper rating except D & E. I see a stronger correlation as compared to Estimated Return. To validate my findings, I calculated correlation by Pearson's method. correlation is -0.84, 95% confidence interval. Not the strongest, but strong enough correlation. I also checked the statistics to validate that there is no overlap between the quartile boundary values except for D & E. D 3rd quartile is 0.24 and E 1st quartile is 0.236, so it's crossing each other.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot4}

# Relationship between LenderYield and ProsperRating

## Boxplot between LenderYield and ProsperRating
ggplot(aes(x = ProsperRating_alpha, y = LenderYield), 
       data = loandata_2009) +
  geom_boxplot() +
  scale_y_continuous(lim = c(0, quantile(loandata_2009$LenderYield, 
                                         0.99)))

## Find the correlation using Pearson's method
with(loandata_2009, cor.test(ProsperRating_numeric, LenderYield))

```
I can see some outliers. Like Estimated Loss chart, the quartile boundaries are separated for all the propsper rating. I see a strong correlation. To validate my findings, I calculated correlation by Pearson's method. correlation is -0.95 with 95% confidence interval, implies the relationship as strong as estimated loss.

### Relationship between main features and other features 

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot5}

# Relationship between MeanCreditScore and ProsperRating

## Boxplot between MeanCreditScore and ProsperRating
ggplot(aes(x = ProsperRating_alpha, y = MeanCreditScore), 
       data = loandata_2009) +
  geom_boxplot() +
  scale_y_continuous(lim = c(600, quantile(loandata_2009$MeanCreditScore, 
                                         0.99)))

## Find the correlation using Pearson's method
with(loandata_2009, cor.test(ProsperRating_numeric, MeanCreditScore))

```
I can see some outliers. Like estimated return chart, the first quartile for one of the lower prosper rating might be less than the third quartile of the higher propser rating. I don't see a strong correlation between the two variables. As calculated by Pearson's method, correlation is 0.55. There is some relationship between two variables, but not significant.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot6}

## ProsperRating vs LoanStatus
ggplot(aes(x = ProsperRating_alpha, fill = LoanStatus), 
       data = loandata_2009) +
  geom_histogram()

## Get statistics
by(loandata_2009$LoanStatus, 
   loandata_2009$ProsperRating_alpha, 
   summary)

```
Depicts the breakdown of each prosper rating by loan status. Breakdown looks similar for almost all the prosper ratings. On looking at the statistics, we can see that breakdown is similar for all prosper rating. Loans in Current status are highest in number, followed by Completed, Charged off and Defaulted.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot7}

## ProsperRating vs IncomeRange
ggplot(aes(x = ProsperRating_alpha, fill = IncomeRange), 
       data = loandata_2009) +
  geom_histogram()

## Get statistics
by(loandata_2009$IncomeRange, 
   loandata_2009$ProsperRating_alpha, 
   summary)

```
Depicts the breakdown of each prosper rating by income range. Number of loans for $100,000+ income range decreases as we move from AA to HR, whereas number of loans for $1-24,999 increases. On looking at the statistics, we can see that the trend identified in the histogram is true. For numbers, please check the summary.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot8}

## ProsperRating vs EmploymentStatus
ggplot(aes(x = ProsperRating_alpha, fill = EmploymentStatus), 
       data = loandata_2009) +
  geom_histogram()

## Get statistics
by(loandata_2009$EmploymentStatus, 
   loandata_2009$ProsperRating_alpha, 
   summary)

```
Depicts the breakdown of each prosper rating by employment status. Breakdown looks similar for almost all the prosper ratings. On looking at the statistics, we can see that breakdown is similar for all prosper rating. Employed borrowers have way more number of loans, as compared to non-employed borrowers. Please check the summary.

### Explore relationship between other variables

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot9}
## Scatterplot for DebtToIncomeRatio vs LoanOriginalAmount 
ggplot(aes(x = LoanOriginalAmount, y = DebtToIncomeRatio), 
       data = loandata_2009) +
  geom_point(alpha = 0.2, position = 'jitter') +
  scale_y_continuous(lim = c(0, 2.5)) +
  scale_x_continuous(lim = c(0, 35000), 
                     breaks = seq(0, 35000, 5000)) +
  geom_smooth(method = 'lm', color = 'green')

## Find the correlation using Pearson's method
with(loandata_2009, cor.test(LoanOriginalAmount, DebtToIncomeRatio))

coef(lm(DebtToIncomeRatio ~ LoanOriginalAmount, data = loandata_2009))

```
Here I am trying to find the relationship between other variables. I am trying to explore how the debt to income ratio affects the amount requested for loan. Due to many overlapping values, I added the regression line to identify the relationship by fitting a linear equation. The regression line is almost parallel to x-axis depicting that the debt to income ratio is similar for all the loan amounts. As per statistics, there is no correlation between two variables. Correlation value is -0.017 with 95% confidence interval. Slope is close to 0, and intercept is 2.7.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
- Here I have explored the relationship of ProsperRating with EstimatedLoss, EstimatedReturn, EstimatedEffectiveYield, and LenderYield. I can see the strongest correlation between prosper rating, estimated loss and lender yield. There is a strong relationship between prosper rating and estimated effective yield. but it is not as strong for estimated return.

- I also investigated the relationship of prosper rating with other categorical features (Loan Status, Income Range, and Employment Status). The breakdown for categorical features looks similar for Loan Status and Employment Status. However, there is a trend for Income Range as explained in the chart. 

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
- Yes, I investiagted if there is a relationship between LoanOriginalAmount and DebtToIncomeRatio. But, as explained above in the chart, I was not able to find any relationship between two variables. 

### What was the strongest relationship you found?
The strongest relationship was found between ProsperRating and EstimatedLoss. It has a negative correlation of -0.96. As the estimated loss increases, prosper rating becomes worse.


# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12, Multivariate_Plot1}

# Plot ProsperRating, color it by IncomeRange and facet it by LoanStatus
ggplot(aes(x = ProsperRating_alpha, fill = IncomeRange), 
       data = loandata_2009) +
  geom_bar(position = 'fill') +
  facet_wrap(~ LoanStatus)

```
This charge depicts the relationship between income range and prosper rating, by loan status. As concluded in above sections, completed, current, charged off, and defaulted holds the most number of loans. These loan statuses have similar distribution of income range, with the highest income range decreasing from AA to HR. However, there are few variations in the pattern for other statuses, the most peculiar being for Past Due (>120 days).There are no highest income range ($100,000+) borrowers or top two prosper rating (AA, A) borrowers with the loan status past dues (>120 days).

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12, Multivariate_Plot2}

# Plot ProsperRating, color it by LoanStatus and facet it by EmploymentStatus
ggplot(aes(x = ProsperRating_alpha, fill = LoanStatus), 
       data = loandata_2009) +
  geom_histogram(position = 'fill') +
  facet_wrap(~ EmploymentStatus)

```
This chart depicts the relationship between loan status and prosper rating, by employment status. For each employment status, we can see the breakdown of loan status over propsper rating. The number of completed loans are more than the current loans for part-time, retired, and full-time employee. However, for employed, self-employed, and other, current are more than the completed. Also, current and completed statuses holds a major percentage of total loans. 

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12, Multivariate_Plot3}

# Plot ProsperRating, color it by IncomeRange and facet it by EmploymentStatus
ggplot(aes(x = ProsperRating_alpha, fill = IncomeRange), 
       data = loandata_2009) +
  geom_histogram() +
  facet_wrap(~ EmploymentStatus, scales = 'free_y')

```
This chart depicts the relationship between income range and prosper rating, by employment status. For each employment status, we can see the breakdown of income range by prosper rating. Most of the part-time and retired borrowers have an income range of 74,999 or less and very few to none with an income of 75,000 or more. We can also see that only for not employed, number of borrowers with HR rating is more than any other ratings. Full-time employees have more borrowers with A rating. 

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12, Multivariate_Plot4}

## Get the sample of 10000 loans from the data set 
set.seed(20022012)

loandata_2009_sample <- loandata_2009[sample(1:length(loandata_2009$ProsperRating_alpha), 10000),
                                      c("ProsperRating_numeric","LenderYield",
                                        "EstimatedEffectiveYield",
                                        "EstimatedLoss", "EstimatedReturn"
                                        )]

colnames(loandata_2009_sample)[colnames(loandata_2009_sample)
                               =="ProsperRating_numeric"] <- "Prosper_Rating"
colnames(loandata_2009_sample)[colnames(loandata_2009_sample)
                               =="LenderYield"] <- "Lender_Yield"
colnames(loandata_2009_sample)[colnames(loandata_2009_sample)
                               =="EstimatedLoss"] <- "Est_Loss"
colnames(loandata_2009_sample)[colnames(loandata_2009_sample)
                               =="EstimatedReturn"] <- "Est_Return"
colnames(loandata_2009_sample)[colnames(loandata_2009_sample)
                               =="EstimatedEffectiveYield"] <- "Est_Yield"

ggpairs(loandata_2009_sample, 
        params = c(shape = I('.'), 
                   outlier.shape = I('.'))) +
  theme(axis.text = element_blank())

```
I am using scatterplot matrices to display the pairwise relationship between my main features. We can see that all the main features are highly correlated to each other, except estimated return.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

- Plotted prosper rating in combination of categorical variables to identify the trends. I don't see that other features strengthen each other in terms of prosper rating. However, I did identified few trends explained below each chart. 

- All the main features are highly correlated to each other, except estimated return. 

### Were there any interesting or surprising interactions between features?
Created a matrix of plots with few variables from the dateset. I could find correlation between ProsperRating_numeric, LenderYield, EstimatedEffectiveYield, EstimatedLoss, EstimatedReturn, ProsperScore. However, all the variables except Lender Yield are estimated/calculated by prosper marketplace, and therefore it is highly likely that they are correlated to each other. To build a predicted model for prosper rating, further investigation is required.  

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12, Plot_One}


loandata_2009_groups <- group_by(loandata_2009, 
                                 ProsperRating_alpha)

loandata_2009_by_prosperrating <- summarise(loandata_2009_groups,
                                            mean_LenderYield = mean(LenderYield),
                                            mean_EstimatedEffectiveYield =
                                              mean(EstimatedEffectiveYield),
                                            mean_EstimatedLoss =
                                              mean(EstimatedLoss),
                                            mean_EstimatedReturn =
                                              mean(EstimatedReturn),
                                            n = n())

p1 = ggplot(aes(x = ProsperRating_alpha, y = mean_EstimatedLoss), 
            data = loandata_2009_by_prosperrating) +
  geom_bar(stat = 'identity', fill = 'orange') +
  ylab("Estimated Loss (Mean)") +
  xlab("Prosper Rating") +
  geom_text(aes(label=round(mean_EstimatedLoss, 2)), vjust = -0.2)
  
p2 = ggplot(aes(x = ProsperRating_alpha, y = mean_EstimatedReturn), 
            data = loandata_2009_by_prosperrating) +
  geom_bar(stat = 'identity', fill = 'orange') + 
  ylab("Estimated Return (Mean)") +
  xlab("Prosper Rating") +
  geom_text(aes(label=round(mean_EstimatedReturn, 2)), vjust = -0.2)

p3 = ggplot(aes(x = ProsperRating_alpha, y = mean_EstimatedEffectiveYield), 
            data = loandata_2009_by_prosperrating) +
  geom_bar(stat = 'identity', fill = 'orange') +
  ylab("Estimated Effective Yield (Mean)") + 
  xlab("Prosper Rating") +
  geom_text(aes(label=round(mean_EstimatedEffectiveYield, 2)), vjust = -0.2)

p4 = ggplot(aes(x = ProsperRating_alpha, y = mean_LenderYield), 
            data = loandata_2009_by_prosperrating) +
  geom_bar(stat = 'identity', fill = 'orange') +
  xlab("Prosper Rating") +
  ylab("Lender Yield (Mean)") +
  geom_text(aes(label=round(mean_LenderYield, 2)), vjust = -0.2)

grid.arrange(p1, p2, p3, p4, ncol = 2, top = "Mean estimations by Prosper Rating")

```

### Description One
As a lender/investor coming to the prosper marketplace, I would like to know few basic insights to make informed decisions. 

As already explored in the above sections, Estimated Loss, Estimated Return, Estimated Effective Yield, and Lender Yield are correlated to prosper rating. In the above plot, I have grouped the records by prosper rating, and then calculated the mean Estimated Loss, Estimated Return, Estimated Effective Yield, and Lender Yield. I have then plotted all the mean values by prosper rating in multiple charts, and presented in a single column grid. 

This visualization will help the lender to get a quick view of the loss/profit rates for each prosper rating, and this will help the lender to make informed decisions on how to invest his money. The genreral trend is that high estimated loss are marked as high risk. In return, high risk loans have high estimated returns and high yield. However, there is an exception where Prosper rating E has comparatively less loss rate than HR, but high return rate. 


### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12, Plot_Two}

final_plot_theme <- theme_grey() + 
  theme(plot.title = element_text(colour ="blue", size = 18), 
        text = element_text(size=12))

ggplot(aes(x = LoanOriginationQuarter), 
       data = subset(loandata_2009, !is.na(LoanOriginationQuarter))) + 
  geom_freqpoly(aes(group = 1)) +
  scale_y_continuous(breaks = seq(0, 30000, 5000)) + 
  final_plot_theme +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust= 1)) +
  xlab("Loan Origination Quarter") +
  ylab("Number of loans (Count)") +
  ggtitle('Frequency plot for number of loans')

loandata_2009_quarters <- group_by(loandata_2009, 
                                 LoanOriginationQuarter)

loandata_2009_by_quarters <- summarise(loandata_2009_quarters,
                                       n = n())


loandata_2009_by_quarters

```

### Description Two
As a lender/investor, I would like to know the frequency of prosper loans. How popular are prosper loans? Is it trending with time or not? 

The frequency plot for total number of loans depicts that prosper loans are becoming popular with time. There is a steep rise from Q1 2013 to Q4 2013. From Q3 2009 to Q1 2011, thee was insignificant growth. However, after that prosper loans started trending and there was tremendous grwoth from Q1 2011 to Q2 2012. There was a little fall from Q2 2012 to Q1 2013, but it picked up again. Propser loansa re rapidly increasing in numbers since Q1 2013 till Q1 2014. I have provided the supporting statistics above. 


### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12, Plot_Three}

ggplot(aes(x = ProsperRating_alpha, fill = IncomeRange), 
       data = loandata_2009) +
  geom_histogram() +
  facet_wrap(~ EmploymentStatus, scales = 'free_y') +
  final_plot_theme +
  guides(fill = guide_legend(title = "Income Range", reverse = TRUE)) + 
  scale_color_brewer(type = 'qual') +
  xlab("Prosper Rating") +
  ylab("Number of loans (count)") + 
  ggtitle('Loans Breakdown by Prosper Rating, Income Range, and Employment Status')

```

### Description Three
As a lender/investor, I would like to view the historical data for the loans by employment status and income range.

With the help of this visualization, I am trying to surface the breakdown ratio of the loan status by prosper rating. I have faceted the plot by employment status to view individual breakdown. This chart depicts the relationship between income range and prosper rating, by employment status. For each employment status, we can see the breakdown of income range by prosper rating. Most of the part-time and retired borrowers have an income range of 74,999 or less and very few to none with an income of 75,000 or more. We can also see that only for not employed, number of borrowers with HR rating is more than any other ratings. Full-time employees have more borrowers with A rating. 

------

# Reflection

### Where did I run into difficulties in the analysis? Where did I find my successes?
It took me a while to understand the data. At the beginning, I wans't sure what am I looking at, what are the key features, and what does each variable means. To better understand the data, I reasearched on internet and read the wiki for prosper loans. I also found the variable definition document for prosper loan dataset, shared by Chris on google drive. The context helped me understand the theoretical relationship between the variables and set the questions for my analysis work. 

I was trying to find some relationship or pattern between prosper rating and categorical variables like income range, employment status. But, I did not find any useful relationship that I could use to predict the prosper rating for new loan listings. However, I did find the strongest relationship between prosper rating and estimated loss rate. Then again, I think estimated loss is predicted by prosper marketplace as well, and it needs further investigation to build a predicted model based on the information provided by the borrower.  

### How could the analysis be enriched in future work (e.g. additional data and analyses)?

As mentioned above, further exploration is required to develop a model to predict the prosper rating based on the information provided by the borrower at the time of loan listing. Also, I do not completely understand all the 81 variables in the dataset, therefore a little more effort is required to study and investigate other variables. 